Robustness of MinD oscillation in \emph{Escherichia coli} with diverse cell shapes

### 2014-09-18 paper/paper.tex
edits to paper after printed out read through Tested everything on quipu.

\section{Introduction}
\fixme{Huang and those who reference him use the terms 'numerical
  deterministic reaction-diffusion model' and 'stochastic version of
  this model' for the stochastic.} It is vital that during the process
of bacterial cell division a cell avoid minicelling, or splitting into
daughter cells with lopsided volumes.  Instrumental to this process in
\emph{Escherichia coli} is a long FtsZ polymer chain that develops on
the cell wall in the center region of the cell, helping dictate the
plane of
division~\cite{adams2009bacterial,lutkenhaus2007assembly}. Previous
experimental studies have shown that the MinC protein, known to
inhibit the FtZ polymer\cite{shen2010examination}, exhibits pole to
pole oscillatory behavior between both ends of the wild-type
pill-shaped cell.  It thus has a higher time averaged concentration in
the cell poles than in the center region, which aides in prohibiting
the FtZ from developing in the wrong region.  The MinC is recruited to
these poles by MinD, which ineracts with another protein, MinE, that
is neccesary for MinD oscillation.~\cite{hu1999topological,
  fu2001mine, shapiro2009and, yu1999ftsz, raskin1999rapid}.

Previous studies have shown that the minD protein system is capable of
exhibiting oscillations in round shapes as well as in connected three
pronged tube shapes~\cite{varma2008min}, in which the oscillations
seem to seek out the extreme poles in the
cell~\cite{corbin2002exploring} \cite{juarez2010changes}. However,
Mannik \emph{et al.} have also shown that there are limitations to
this capability~\cite{mannik2010bacteria}
\cite{mannik2009bacterial}. They have experimentally forced E. Coli
cells into flattened irregular shapes of $0.25\mu m$
thickness. Oscillations in this type of cell shape lose their
regularity, both temporally and spatially. This allows for an
opportunity to test previously studied MinD simulation models against
more extreme experimental cases than have been seen thus far.

A significant amount of work has been done to develop protein reaction
and diffusion models that exhibit accurate macroscopic dynamics of the
MinD protein system. Early models involve free proteins that affect
each others' rates of diffusion and membrane attachment, but do not
combine into compound states~\cite{meinhardt2001pattern}.  In 2003
Huang improved upon this work with a simple and very successful
simulation model based on MinD-MinE combination, ATPase hydrolysis,
and MinD membrane attachment that exhibits accurate MinD oscillations
in cylindrical cells~\cite{huang2003dynamic}. In this model
cytoplasmic MinD is more likely to attach to the membrane when MinD is
already clustered there (following observed non-linear attachement of
minD on the cell membrane), and is stationary once attached.  A number
of studies have used an approach similar in that they do not rely on
the ability of MinD to move along the walls and
cluster~\cite{kruse2007experimentalist, meinhardt2001pattern,
  drew2005polymerization, fange2006noise, kerr2006division}, while
studies have been made as well of models which rely on MinD mobility
and attraction on the cell membrane~\cite{kruse2002dynamic,
  howard2005cellular}.  Experimental studies of the Min system's
association with the cell membrane have also been
made~\cite{hsieh2010direct,mileykovskaya2003effects}.

Variations of the Huang 2003 model that stochasticly account for
variations of molecular
interaction~\cite{kerr2006division,fange2006noise} and monte-carlo
simulations that implement this stochastic version of Huang's mean
field reaction rates confirm the major results obtained by Huang's
model, and more successfully predict experimentally observed
oscillations in round cell phenotypes~\cite{fange2006noise,
  huang2004min}. The successful simulations of three pronged cells
mentioned above are stochastic
simulations~\cite{varma2008min}. Biochemical models of broader scope
have also been used to study the MinD system and show consistent
results~\cite{arjunan2010new}.

@@

where $\rho$ is cytoplasmic protein density ($\mu m^{-3}$), $\sigma$
is membrane bound density ($\mu m^{-2}$), $\mathcal{D}_D$ and
$\mathcal{D}_{E}$ th diffusions constants for MinD and MinE,
respectively, $k_D^{\textrm{ADP $\rightarrow$ ATP}}$ the rate of
conversion from MinD:ADP to the MinD:ATP complex, $k_D$ the rate of
MinD:ATP attachement to the membrane when no protein is already
attached there, $k_{dD}$ the increase of this rate when MinD:ATP is
present on the membrane, $k_{de}$ the rate of hydrolisis of the
MinD:MinE:ATP complex, $k_E$ the rate of cytoplasmic MinE binding to
membrane bound MinD:ATP complex, and $d_w$ is the distance from the
point in space to the closest wall (which will always be
perpendicular to this distance).  $\delta(d_w)$ has units of
$\mu m^{-1}$ and will be zero everywhere except at the wall.
Equations \ref{eq:d-on-wall} and \ref{eq:FifthPDE} will only be
relevent at the membrane because the membrane bound density values
will have no meaning in the cytoplasm where there is no membrane.

Our diffusion and reaction rates are shown below.  We are interested
primarily in the effect of cellular size and shape on the protein
oscillations, so we follow Huang\cite{huang2003dynamic} and do not
deviate from the wild type values used in the cited work (see values
below).  Huang's simulations use total MinD and MinE concentrations of
$1,000/\mu m$ and $350/\mu m$, respectively, in a cylindrical cell of
radius $0.5\mu m$, and in our (non-cylindrical) cells we use the same
number of proteins per unit volume.  These concentration values are
1273 $\mu m^{-3}$ and 446 $\mu m^{-3}$, respectively. \fixme{It's been
  a while so might want to check this to make sure it's true.}
\begin{gather*}
  \mathcal{D}_D = \mathcal{D}_{E} = 2.5\micron^2/\text{sec}\\
  k_D^{\textrm{ADP $\rightarrow$ ATP}} = 1/\textrm{sec,  }
  k_D = 0.025 \micron /\textrm{sec}\\
  k_{dD} = 0.0015 \micron^3/ \textrm{sec,  }
  k_{de} = 0.7/\textrm{sec}\\
  k_E = 0.093 \micron^3 /\textrm{sec}.
\end{gather*}
A 3D grid is then constructed in cartesian coordinates, with a grid
spacing of .05\micron.

We have performed both a basic finite analysis, deterministic
simulation that is spatially and temporally discrete, and a stochastic
analysis that is spatially discrete but continuous in time.  Our
stochastic simulation method follows the work of
Kraus~\cite{kraus1996crosstalk} which in turn follows a method
introduced by Gillespie~\cite{gillespie1977exact}.

We mean to investigate the geometric limits of the Min system
oscillations as observed by Mannik \emph{et
  al.}~\cite{mannik2012robustness}, so we model the Min system in a
variety of cell shapes and sizes.  Here we present a selection of
these, first naturally occuring pill-shaped cells and then a number of
flattened out shapes which reflect the experiments of Mannik \emph{et
  al.}, in which bacteria are confined within a thin slit of height
$0.25\micron$ ~\cite{mannik2012robustness}. Viewed from the top down
the cells will show the shapes described below and viewed from the
side they show at their edges a semicircular protrusion (one may
imagine the shape of a pancake). We have simulated and analyzed a
number of these flattened shapes in addition to those shown below,
including equilateral and iscosolese triangle shapes, shapes shaped
like greater than signs, '$>$', and four-pronged 'starfish' shapes.
For all of these we have varied the cell size, expanding the two
dimensions that are tangent to the flattened surface from very small,
'steady state' scales (which we discuss below) to scales twice that of
the published Mannik shapes.  Our conclusions are based on all of this
data, but we present here specific, illustrative examples: two
flattened shapes that are designed to replicate those published by
Mannik and two `stadium' shapes that respectively have the same
quadrupole moments of the two Mannik shapes.

@@

Huang \emph{et al} ~\cite{huang2003dynamic} has has performed a linear
stability analysis on the model which shows an upper limit on a steady
state solution of $2\mu m$.  Cells with dimensions longer than this
are unstable and show oscillatory behavoir in this direction, where
cells that are shorter relax into a motionless steady state.  We have
similarly performed a stability analysis in an infinite slab with a
thickness equal to our flattened cells ($0.25\mu m$) and have found
that an equivalent stability limit for half wavelengths is $2.13\mu
m$. As expected, when decreasing the lengths and widths of our
simulated flattened cells so that the shortest distance across the
cell is less than this length, the cells stop exhibiting any
oscillalitory behavior.  The deterministic simulations relax into a
motionless state and the stochastic simulations exhibit multiple
random locations of small maximization.

@@

In contrast, the stochastic method~\fixme{cite} predicts spatially
irregular formation of maxima for all the flattened cells that we
studied.  In order to better understand the irregular oscillations
observed in experiment, we simulated flattened ``stadium'' shapes in
order to distinguish between the effect of flattening and the effect
of irregular shape.  We found that although the stadium cells
exhibited somewhat more irregular locations of maxima than the
wild-type cells, their oscillation was far more periodic and bipolar
than the irregular cell shapes observed experimentally.  Thus we
confirm that the disruption of the MinD oscillations seen by Mannik
\emph{et al.} is largely due to the irregularity of cell shapes
observed.

>>>>

\section{Introduction}
\fixme{Huang and those who reference him use the terms 'numerical
  deterministic reaction-diffusion model' and 'stochastic version of
  this model' for the stochastic.} It is vital that during the process
of bacterial cell division a cell avoid minicelling, or splitting into
daughter cells with lopsided volumes.  Instrumental to this process in
\emph{Escherichia coli} is a long FtsZ polymer chain that develops on
the cell wall in the center region of the cell, helping dictate the
plane of
division~\cite{adams2009bacterial,lutkenhaus2007assembly}. Previous
experimental studies have shown that the MinC protein, known to
inhibit the FtZ polymer\cite{shen2010examination}, exhibits regular
pole to pole oscillatory behavior between both ends of the wild-type
pill-shaped cell.  It thus has a higher time averaged concentration in
the cell poles than in the center region, which aides in prohibiting
the FtZ from developing in the wrong region.  The MinC is recruited to
these poles by MinD, which ineracts with another protein, MinE, that
is neccesary for MinD oscillation.~\cite{hu1999topological,
  fu2001mine, shapiro2009and, yu1999ftsz, raskin1999rapid}.

Previous studies have shown that the minD protein system is capable of
exhibiting oscillations in round shapes \fixme{citation needed}as well
as in connected three pronged tube shapes~\cite{varma2008min}, in
which the oscillations seem to seek out the extreme poles in the
cell~\cite{corbin2002exploring} \cite{juarez2010changes}. However,
Mannik \emph{et al.} have also shown that there are limitations to
this capability~\cite{mannik2010bacteria}
\cite{mannik2009bacterial}. They have experimentally forced E. Coli
cells into flattened irregular shapes of $0.25\mu m$
thickness. Oscillations in this type of cell shape lose their
regularity, both temporally and spatially. This allows for an
opportunity to test previously studied MinD simulation models against
more extreme experimental cases than have been seen thus
far. \fixme{Perhaps this sentence should come later after we've
  introduced the simulation research that's been done.}

A significant amount of work has been done to develop protein reaction
and diffusion simulation models that exhibit accurate macroscopic
dynamics of the MinD protein system. Early models involve free
proteins that affect each others' rates of diffusion and membrane
attachment, but do not combine into compound
states~\cite{meinhardt2001pattern}.  In 2003 Huang improved upon this
work with a simple and very successful simulation model based on
MinD-MinE combination, ATPase hydrolysis, and MinD membrane attachment
that exhibits accurate MinD oscillations in cylindrical
cells~\cite{huang2003dynamic}. In this model cytoplasmic MinD is more
likely to attach to the membrane when MinD is already clustered there
(following observed non-linear attachement of minD on the cell
membrane), and is stationary once attached.  A number of studies have
used an approach similar in that they do not rely on the ability of
MinD to move along the walls and
cluster~\cite{kruse2007experimentalist, meinhardt2001pattern,
  drew2005polymerization, fange2006noise, kerr2006division}, while
studies have been made as well of models which rely on MinD mobility
and attraction on the cell membrane~\cite{kruse2002dynamic,
  howard2005cellular}.  Experimental studies of the Min system's
association with the cell membrane have also been
made~\cite{hsieh2010direct,mileykovskaya2003effects}.

Variations of the Huang 2003 model that stochasticly account for
molecular interaction and
diffusion~\cite{kerr2006division,fange2006noise} and monte-carlo
simulations that implement this stochastic version of Huang's mean
field reaction rates \fixme{citation needed on the monte-carlo method}
confirm the major results obtained by Huang's model, and more
successfully predict experimentally observed oscillations in round
cell phenotypes~\cite{fange2006noise, huang2004min}. The successful
simulations of three pronged cells mentioned above are stochastic
simulations~\cite{varma2008min}. Biochemical models of broader scope
have also been used to study the MinD system and show consistent
results~\cite{arjunan2010new}.

@@

where $\rho$ is cytoplasmic protein density ($\mu m^{-3}$), $\sigma$
is membrane bound density ($\mu m^{-2}$), $\mathcal{D}_D$ and
$\mathcal{D}_{E}$ th diffusions constants for MinD and MinE,
respectively, $k_D^{\textrm{ADP $\rightarrow$ ATP}}$ the rate of
conversion from MinD:ADP to the MinD:ATP complex, $k_D$ the rate of
MinD:ATP attachement to the membrane when no protein is already
attached there, $k_{dD}$ the increase of this rate when MinD:ATP is
present on the membrane, $k_{de}$ the rate of hydrolisis of the
MinD:MinE:ATP complex, $k_E$ the rate of cytoplasmic MinE binding to
membrane bound MinD:ATP complex, and $d_w$ is the distance from the
point in space to the closest wall (which will always be
perpendicular to this distance).  $\delta(d_w)$ has units of
$\mu m^{-1}$ and will be zero everywhere except at the wall.
Equations \ref{eq:d-on-wall} and \ref{eq:FifthPDE} will only be
relevent at the membrane because the membrane bound density values
will have no meaning in the cytoplasm.

Our diffusion and reaction rates are shown below.  We are interested
primarily in the effect of cellular size and shape on the protein
oscillations, so we follow Huang\cite{huang2003dynamic} and do not
deviate from the wild type values used in the cited work.

\begin{gather*}
  \mathcal{D}_D = \mathcal{D}_{E} = 2.5\micron^2/\text{sec}\\
  k_D^{\textrm{ADP $\rightarrow$ ATP}} = 1/\textrm{sec,  }
  k_D = 0.025 \micron /\textrm{sec}\\
  k_{dD} = 0.0015 \micron^3/ \textrm{sec,  }
  k_{de} = 0.7/\textrm{sec}\\
  k_E = 0.093 \micron^3 /\textrm{sec}.
\end{gather*}

Huang's simulations use total MinD and MinE concentrations of
$1,000/\mu m$ and $350/\mu m$, respectively, in a cylindrical cell of
radius $0.5\mu m$, and in our (non-cylindrical) cells we use the same
number of proteins per unit volume.  These concentration values are
1273 $\mu m^{-3}$ and 446 $\mu m^{-3}$, respectively. \fixme{It's been
  a while so might want to check this to make sure it's true.}  A 3D
grid is then constructed in cartesian coordinates, with a grid spacing
of .05\micron.

We have performed both a basic finite analysis, numerical,
deterministic simulation that is spatially and temporally discrete,
and a stochastic analysis that is spatially discrete but continuous in
time.  Our stochastic simulation method follows the work of
Kraus~\cite{kraus1996crosstalk} which in turn follows a method
introduced by Gillespie~\cite{gillespie1977exact}.

We mean to investigate the geometric limits of the Min system
oscillations as observed by Mannik \emph{et
  al.}~\cite{mannik2012robustness}, so we model the Min system in a
variety of cell shapes and sizes.  Here we present a selection of
these, first naturally occuring pill-shaped cells and then a number of
flattened out shapes which reflect the experiments of Mannik \emph{et
  al.}, in which bacteria are confined within a thin slit of height
$0.25\micron$ ~\cite{mannik2012robustness}. Viewed from the top down
the cells will show the shapes described below and viewed from the
side they show at their edges a semicircular protrusion (one may
imagine the shape of a pancake). We have simulated and analyzed a
number of these flattened shapes in addition to those shown below,
including equilateral and iscosolese triangle shapes, shapes shaped
like greater than signs, '$>$', and four-pronged 'starfish' shapes.
For all of these we have varied the cell size, expanding the two
dimensions that are tangent to the flattened surface from very small,
'steady state' scales (which we discuss below) to scales twice that of
the published Mannik shapes.  Our conclusions are based on all of this
data, but we present here specific, illustrative examples: two
flattened shapes that are designed to replicate those published by
Mannik and two `stadium' shapes that respectively have the same aspect
ration, thickness, and volume as the two Mannik shapes.  Viewed from
the top down, the stadium shapes appear as rectangles with
semi-circular end caps on their long axis ends.

@@

Huang \emph{et al} ~\cite{huang2003dynamic} has performed a linear
stability analysis on the model which shows an upper limit on a steady
state solution of a $2\mu m$ wavelength.  Cells with dimensions longer
than this are unstable and show oscillatory behavoir in this
direction, while cells that are shorter relax into a motionless steady
state.  We have similarly performed a stability analysis in an
infinite slab with a thickness equal to that of our flattened cells'
($0.25\mu m$) and have found that an equivalent stability limit for
half wavelengths is $2.13\mu m$. As expected, when decreasing the
lengths and widths of our simulated flattened cells so that the
longest distance across the cell is less than this length, the cells
stop exhibiting any oscillalitory behavior.  The deterministic
simulations relax into a motionless state and the stochastic
simulations exhibit multiple random locations of small maximization.

@@

As found in previous studies~\fixme{cite}, we observe that the
stochastic method predicts behavior in wild type cells that is
consistent both with experiment and with the deterministic method.  In
contrast, the stochastic method~\fixme{cite} predicts spatially
irregular formation of maxima for all the flattened cells that we
studied.  In order to better understand the irregular oscillations
observed in experiment, we simulated flattened ``stadium'' shapes in
order to distinguish between the effect of flattening and the effect
of irregular shape.  We found that although the stadium cells
exhibited somewhat more irregular locations of maxima than the
wild-type cells, their oscillation was far more periodic and bipolar
than the irregular cell shapes observed experimentally.  Thus we
confirm that the disruption of the MinD oscillations seen by Mannik
\emph{et al.} is largely due to the irregularity of cell shapes
observed, as opposed to the size or aspect ratio.
